\documentclass[11pt]{amsart}
\usepackage{amssymb, amsmath}



\title[Hi]{Nice title}
\author{Arthur Writer}
\address{Department of Mathematics, Some University}
\email{awriter@some.edu}
\date{\today}




\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\KK}{\mathbb{K}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand{\Ocal}{\mathcal{O}}
\newcommand{\Es}{\mathcal{E}}
\newcommand{\Ncal}{\mathcal{N}}
\newcommand{\basin}{\mathcal{B}}
\newcommand{\Supp}{\operatorname{Supp}}
\newcommand{\Spec}{\operatorname{Spec}}
\newcommand{\Gal}{\operatorname{Gal}}
\newcommand{\PSL}{\operatorname{PSL}}
\newcommand{\GL}{\operatorname{GL}}
\newcommand{\Aut}{\operatorname{Aut}}
\newcommand{\ord}{\operatorname{ord}}
\newcommand{\MOD}[1]{~(\textup{mod}~#1)}
\newcommand{\h}{\hat{h}}
\newcommand{\im}{\operatorname{im}}
\newcommand{\Pic}{\operatorname{Pic}}

\newcommand{\affine}{\mathbb{G}_{\operatorname{Aff}}}

\newcommand{\pf}{\mathfrak{p}}
\newcommand{\mf}{\mathfrak{m}}
\renewcommand{\div}{\operatorname{div}}
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\greens}{\mathcal{G}}

\newcommand{\llbracket}{[\hspace{-1.5pt}[}
\newcommand{\rrbracket}{]\hspace{-1.5pt}]}
\newcommand{\llp}{(\hspace{-1.5pt}(}
\newcommand{\rrp}{)\hspace{-1.5pt})}



\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{remark}{Remark}



\begin{document}
\begin{abstract}
I'm an abstract
\end{abstract}


\maketitle

I want to talk about things but notice that this part does not belong to a section!

 \begin{theorem}\label{th:some_theorem}
I'm a theorem.
\end{theorem}

Blah blah blah

\begin{theorem}\label{th:some_other_theorem}
Impart me with mathematical knowledge!
\end{theorem}

Maybe a corollary

\begin{corollary}\label{cor:a_corollary}
I immediately follow from the above theorem.
\end{corollary}


More stuff!

\begin{corollary}\label{cor:another_corollary}
More delicious mathematical knowledge.
\end{corollary}

Maybe you could describe how we demonstrate this corollary.



\section{Proof of Theorem~\ref{th:main}}

Now this is finally in a section.

\begin{lemma}
Some lemma
\end{lemma}


\begin{proof}
Maximum effort!
\end{proof}

Blah blah blah.

\subsection{This is a subsection}

I'm about one thing.

\subsection{This is another subsection}

I'm about another thing.

\end{document}
